---
title: "Merging CSV files"
output:
  html_notebook:
    code_folding: show
    toc: yes
    theme: cerulean
  html_document:
    df_print: paged
    toc: yes
  pdf_document: default
---

<style type="text/css">

body{ /* Normal  */
      font-size: 14px;
  }
td {  /* Table  */
  font-size: 14px;
}
h1.title {
  font-size: 38px;
}
h1 { /* Header 1 */
  font-size: 28px;
}
h2 { /* Header 2 */
    font-size: 22px;
}
h3 { /* Header 3 */
  font-size: 18px;
}
code.r{ /* Code block */
    font-size: 14px;
}
pre { /* Code block - determines code spacing between lines */
    font-size: 14px;
}
</style>

# Problem
We have a few CSV files that have the same columns. We want to merge them together while retaining information about which file each row came from. 

# Data
Let's say the CSV files are `file{1,2,3,4}.csv`. This is `file1.csv`:

```{r message=FALSE}
library(tidyverse)
read_csv("file1.csv")
```
The other three files are similar. 

# Solution summary

This is the complete solution using tidyverse libraries:

```{r message=FALSE}
library(tidyverse)
csv_names = paste0("file", 1:4, ".csv") 
csv_names %>% map(read_csv) %>% set_names(csv_names) %>% bind_rows(.id = "file")
```

# Breaking it down 

## Making the list of files 

The first step is to get all the file names in a list. If your filenames have a regular pattern, you can construct that easily with `paste`. `paste0` is a special case of paste with the joining delimiter set to `""`. 

```{r}
csv_names = paste0("file", 1:4, ".csv") 
csv_names
```

## Import each of them 

Now we want to import each of these files into a data frame (technically a tibble since we're using tidyverse). Fortunately `purr` has a map function that works very similarly to map function in Javascript, which I therefore found much more intuitive to use than the more popular `lapply` variety of functions in R.  

```{r message=FALSE}
csv_dfs = csv_names %>% map(read_csv)

length(csv_dfs)
csv_dfs[[3]]
```

## Joining them together

The [`bind_rows`](https://dplyr.tidyverse.org/reference/bind.html) function binds rows together, and has a handy `.id` argument that creates a new column holding data that links back the rows to the data frames they came from. 

```{r rows.print=8}
csv_dfs %>% bind_rows(.id="filename")
```

This is still not quite exactly what we want, because unless we name the elements in our list, bind_rows does not does not know what to put in the `.id` column, so it just uses a numeric sequence instead. `lists` in R are almost like dictionaries in python, but some elements only has a value and no key. Right now our list has only values and no keys. Setting the name for all of them can be done in one shot using `set_names`.

```{r message=FALSE}
csv_dfs = csv_names %>% map(read_csv) %>% set_names(csv_names)

names(csv_dfs)
csv_dfs$file3.csv
```

Now `bind_rows` can use these names to set the values in the `.id` column: 

```{r}
csv_dfs %>% bind_rows(.id="filename")
```


We're done! 

##  Modifying each csv before merging

You may want to modify each individual CSV before merging. For example, your original csv files may have 100s of columns but you only want to keep a few, and you don't want to merge first and then select for efficiency. It is easy to do so by creating a wrapper function around `read_csv`: 

```{r message=FALSE}

csv_extract = function(csv_name) {
  read_csv(csv_name) %>% 
    select(Red, Blue, Green)  # Dropping Yellow because we're not interested.
}
csv_names = paste0("file", 1:4, ".csv") 
csv_names %>% map(csv_extract) %>% set_names(csv_names) %>% bind_rows(.id = "file")
```
